Spatial and temporal variability in economic availability of residual biomass

### 2021-11-19 analysis/manuscript.Rmd
Jay's intro'

---
title: Spatial and temporal variability in economic availability of residual biomass
  from timber harvest in west-central Alberta, Canada (1990--2015)
author:
- Glen W. Armstrong:
    institute: rr
    correspondence: 'yes'
    email: gwa@ualberta.ca
- Jay A. Anderson:
    email: jaya@ualberta.ca
    institute: rees
- Alex W. McPhee:
    email: awmcphee@ualberta.ca
    institute: rees
institute:
- rr: Department of Renewable Resources, University of Alberta, Edmonton AB  T6G 2H1,
    Canada
- rees: Department of Resource Economics and Environmental Sociology, University of
    Alberta, Edmonton AB  T6G 2H1, Canada
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    number_sections: yes
    df_print: paged
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
  word_document:
    number_sections: yes
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
  pdf_document:
    includes:
      in_header: preamble.tex
    number_sections: yes
    latex_engine: xelatex
    keep_tex: yes
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
header-includes:
- \usepackage[left]{lineno}
- \usepackage{setspace}
- \linenumbers
geometry: left=1.5in,right=0.5in,top=1in,bottom=1in
csl: biomass-and-bioenergy
bibliography: historicalResidueSupply.bib
---

\raggedright
<!-- \fontfamily{cmr} -->
<!-- \fontsize{12}{22} -->
<!-- \fontseries{b} -->
<!-- \selectfont -->

@@

```{r libraries, echo=FALSE, message=FALSE}
setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
library(tidyverse)
library(ggthemes)
library(gridExtra)
library(viridis)
library(skimr)
library(scales)
library(DescTools)
library(zoo)
library(ezknitr)
library(magick)
library(kableExtra)
library(gt)
library(glue)
```

\newpage
\doublespacing

@@

<!-- ## Some obvious citations -->

<!-- @Carriquiry2011 -->

<!-- @Dessbesell2017 -->

<!-- @Dymond2010 -->

<!-- @Lo2021 -->

<!-- @Malladi2018 -->

<!-- @Mansuy2017 -->

<!-- @Mansuy2020 -->

<!-- @Schroder2019 -->

<!-- @Shabani2013 -->

<!-- @Shooshtarian2018 -->

<!-- @Yemshanov2018 -->

<!-- @Zheng2021 -->

<!-- @Ackom2010 -->

<!-- @Acuna2019 -->

@@

@@

@@

We used a number of publicly available datasets to develop the datasets
used in our analysis.  Most of the data are freely available on the World
Wide Web (WWW). Much of the data processing required for this study was 
manipulation of spatial
data.  This was done using the QGIS software @QGISorg2021.
All national level data were clipped to the Alberta provincial boundary
for further analysis.

The datasets are only introduced here.  The way they are used will be discussed
in detail later.

@@

Altalis manages much of the spatial data created by or for the Government of
Alberta.  We use several coverages from their base features database product 
@Altalis2021:
the provincial, green/white area, forest management unit (FMU), forest management
agreement area (FMAA), Alberta Township Survey System
township, and town
boundaries.

The Alberta Township Survey System (ATS) is particularly important to this study
as all the data used for this study are aggregated or disaggregated to the
township level.  ATS provides the basis for legal land descriptions in Alberta.
The typical township is roughly square (9.778 km by 9.716 km, 9 500 ha) and 
is indexed by
township (1--126), range (1--30), and meridian (4--6) numbers.  
The townships immediately east of
Alberta's western boundary, and immediately east of the 5^th^ and 6^th^
meridians (corresponding to 114°W and 118°W longitude, respectively) are 
partial townships and therefore deviate from the roughly square shape.  ATS
is described in detail by the Alberta Land Surveyor's Association @ALSA2021.
There are 7 237 townships in Alberta.  That allow for pretty good spatial 
resolution.

@@

Alberta Geospatial Services is operated by the Alberta Ministry of Environment
and Parks @Alberta2021a.  We obtained the road access coverages from here.  The
road access coverages available from Altalis do not have as detailed information
on road class as those available from Alberta Geospatial Services.

@@

As part of Canada's National Forest Inventory, the National Forest Information
System (NFIS) provides a web-based individual tree biomass calculator @NFIS2021. 
It 
is used to calculate stem wood, stem bark, branch, and foliage biomass for 
a tree based on province, ecozone, species, diameter at breast height (dbh) and
height.  It is possible to upload a file representing many trees with different
combinations of these input parameters and download a file containing the
calculated results for all the trees.

NFIS has also produced a set of spatial data layers representing forest 
properties at a 250 m resolution for all of Canada @Beaudoin2017.  We used the
merchantable volume, stand age, and species composition attributes from these 
data layers.

Another product of NFIS is a set of spatial datasets derived from satellite data 
@NFIS2021b.  The dataset used for this study is the harvest year data set for
the years 1985-2015.  The dataset is described in detail by Hermosilla *et al.*
@Hermosilla2016.  Essentially, it consists of a raster for Canada at 30 m 
resolution indicating the year of harvest for each cell.

@@

For our study, the most important product of Natural Resources Canada is
the Ung *et al.* paper @Ung2013 which describes the tree taper models 
developed for Canada.  A valuable contribution related to this is the WWW-based
wood volume calculator that uses those models @NRCan2015.  It is helpful for
understanding how taper functions work.

@@

The national forestry database @CCFM2020 was used to summarize the 
timber harvest volume from
Alberta for the years 1990-2015.

@@

The Alberta government requires FMA holders to develop a forest management
plan for their FMAA.
The forest management plans were used to get information on utilization 
standards and piece size of harvested trees @Alberta2021.

### Miscellaneous

The map of provincial boundaries comes from Statistics Canada @Canada2016.

The boundaries of Canada's terrestrial ecozones, which are important inputs to
the biomass calculator can be found on Agriculture and Agri-Food Canada's
web site @Canada2017.

Alberta is a province in western Canada.  It occupies 661 848 km^2^.
Its southern boundary is 49°N latitude, northern boundary is 60°N latitude,
eastern boundary is 110°W longitude, and its western boundary is defined by
120°W longitude and the Great Divide of the Rocky Mountains.

The southeastern part of the province is largely privately owned agricultural
land, but the northern and western parts are largely publicly owned forest land.
There is a substantial area of privately owned agricultural land
in the Peace River Country of northwestern Alberta.  
There are also substantial areas
in National Parks along the Rocky Mountains and in the far north of the 
province.  The province calls the area of mostly publicly owned forest land 
as the Green Area, and the area of mostly privately owned agricultural land 
as the White Area.  About 90% of the harvested timber volume in Alberta
comes from the Green Area.

For administrative purposes, the Green Area is divided into Forest Management
Units (FMUs).  The province also enters into long term agreements with forestry
companies giving the companies the rights to timber harvest and the 
responsibility to manage the forest sustainably.  These agreements are
called Forest Management Agreements, and the areas under agreement are
called Forest Management Agreement Areas (FMAAs).  For the most part, the
boundaries of the FMAAs are now coincident with FMU boundaries, with
the exception that the Alberta-Pacific FMAA in eastern Alberta comprises
12 FMUs.
<!-- FMAAs occupy 62% of the Green Area **and maybe something about proportion of -->
<!-- harvest** -->

Our focus for this study is the forest near Hinton, Edson, and Whitecourt in 
west-central Alberta.  These towns were chosen as they host several
forest products companies and are located near very active FMAAs.
We assume that it would be likely that a forest residue processing plant
would be located near an existing forest products mill as the forest road
network has been developed to supply those mills, and location of a residue
plant near a forest products mill would also allow for access to 
mill residues.  Figure ?? shows the location of the three towns in 
relation to FMAAs and the Green and White Areas.

The FMAAs which supply most of the timber to mills in our near our three 
locations are West Fraser Mills Ltd. (Hinton), West Fraser Mills Ltd. (Edson),
Weyerhaeuser Company Limited (Pembina Timberlands), Millar Western, Blue
Ridge Lumber Inc., and ANC Timber Ltd.

For this study, we assumed that forest harvest residue processing would take
place at a plant co-located with an existing forest products manufacturing
facility. We used the Weyerhaeuser oriented strand board plant in Edson, the
Millar Western Forest Products Ltd. pulp mill in Whitecourt, and the West Fraser
pulp mill in Hinton as the locations for the hypothetical forest residue
processing facilities.

As well as considering locations for the hypothetical plants, we considered 3
different plant capacities: 200 Gg/a  corresponding to a small wood pellet
mill, 400 Gg/a  corresponding to a large pellet mill or a small ethanol
plant, and 800 Gg/a  corresponding to a large ethanol plant.

The areas harvested in Alberta in 2009 according to NFIS @NFIS2021b
@Hermosilla2016 are shown in relationship to the three towns and 
the Green Area in Figure 3.  Our data set includes harvest areas for the
years 1990--2015, so this is just an example.

Figure 4 shows the proportion of the township area harvested in 2009 aggregated
to the township level, showing only those townships where the proportion of
the township area harvested was greater than 0.001.  For each township, and
each of the years between 1990 and 2015, we store the area harvested (ha) for
use in our analysis.  We can see from this map that a great deal of timber
harvest took place near our plant locations in 2009.

@@

The provincial road network was sorted into four classes: primary highways,
other paved roads, two-lane gravel roads, and one-lane gravel or dirt roads.  
Loaded and unloaded travel speeds were assumed for each road class, according
to the table below. The harmonic mean of loaded and unloaded speeds 
(rounded to the nearest
multiple of 5 km/h) were calculated as
$$
\bar{s_h}= \left(\frac{s_l^{-1} + s_u^{-1}}{2}\right)^{-1}
$$
where $s_l$ is the loaded speed and $s_u$ is the unloaded speed.
Use of harmonic mean allows us to calculate the correct  cycle time:  the
time required to travel from the mill, to the collection site in the forest,
and back to the mill.  This provides the basis for our estimates of haul cost.
Calculating average haul speed using an arithmetic mean leads to a biased
estimate of cycle time.

<!-- Check out https://haozhu233.github.io/kableExtra/awesome_table_in_html.html to -->
<!-- try to format this table better. -->

Figure 6 illustrates the road network in Alberta and estimates of one-way haul
time based on the average of loaded and unloaded speeds from Table ??. The
road network in the White Area of the province is very dense and largely arranged
as a grid related to the Alberta Township Survey System.  Most of the roads
in the White Area are public.  The road network in the Green Area is much
less dense and many of the roads are built and maintained by resource industries 
(*e.g.* forestry and petroleum & natural gas).  The one-way haul times were
generated using the QGIS Network Analysis Toolbox 3 plugin (QNEAT3) created by 
Raffler @Raffler2018,
specifically its iso-area as interpolation (from point) algorithm.  In our case,
we used the algorithm to determine the minimum one-way travel time along the road network 
from each
of our centres to each cell of a provincial level raster at a 200 m resolution.  Travel time
was calculated based on the harmonic mean of loaded and unloaded times for each
road segment.  Off-road travel was assumed to be at 5 km/h.  The underlying
algorithm for travel time minimization is that of Dijkstra @Dijkstra1959.

QGIS zonal statistics were used to calculate the median network travel time to
each township, from each of our three centres.  The one-way travel time in Figure
6 was limited to 6 hours because this equates to a 13 hour cycle time if
you assume 1 hour for loading and unloading.  In Alberta, drivers of commercial
vehicles are not permitted to drive more than 13 hours in a day.  **technically
we could take the picture out to 6.5 hours, as the loading and unloading time
does not count as driving time.  There is an on-duty limit of 14 hours**.

@@

merch_vol_adj_factor <- 2.77 # based on a wild ass guess
#merch_vol_adj_factor <- 2.38 #based on West Fraser Hinton
odmt_adj_factor <-  0.10

@@

```
### Utilization standards

Many forest products companies operate in Alberta, and have different agreements
with the government with respect to how the forest is managed.  Some companies
have harvesting rights to just the needleleaf species, others have rights to
just the broadleaf species, and some have rights to both needleleaf and
broadleaf.

Utilization standards in Alberta are specified on the basis of stump height,
minimum stump diameter (outside bark), and minimum log length to a specified top
diameter (inside bark).  Utilization standards vary by FMU.  Stump heights used
in Alberta are 15 and 30 cm; stump diameters are 13 or 15 cm; top diameters are
7, 10, 11, or 12 cm; and minimum log lengths are 2.44, 2.66, 2.67, 3.66, 3.76, 
or 4.88 m.  These utilization standards are necessary to relate volumes 
calculated to the 13/7 standard for the National Forest Inventory to the harvest
volumes reported by the companies.  Perhaps most importantly, the minimum top
diameter will have an effect on the volume of residue left in the forest 
corresponding to the tops of trees.

@@

All measures of biomass (Mg) used in this study are reported on
a dry matter basis.

We assume that the biomass available for collection at a cutblock includes the
tops, branches, and foliage left at a roadside landing after processing by
a stroke delimber.  The delimber is assumed to top the log at the top diameter
specified in the utilization standard relevant to the harvest area.  Recovery
will not be perfect.  We assume a recovery factor of 0.624 
following Peltola *et al.* 
@Peltola2011, recognizing that some of the residual biomass will 
be unrecoverable.

Our focus in this study is on the spatial and temporal variation in biomass
collection costs based on residue availability associated with cycle times.  Following Röser @Roser2013, the pre-piling cost was set to
`r sprintf("%5.2f",prepiling_cost)` \$·Mg^-1^ , the comminution cost (@Blackburn2020a)
was set to `r sprintf("%5.2f",comminution_cost)`  \$·Mg^-1^ , and other costs
(roads, supervision, overhead, maintenance, and compliance) were
set to  `r sprintf("%5.2f",other_cost)` \$·Mg^-1^ .  The total cost,
excluding transport, is
`r sprintf("%5.2f",constant_cost)` \$·Mg^-1^ .

We assume that the residue will be transported from the forest using a 
live floor chip van with 100 m³ capacity.
The USDA Forest Service's Forest Residue Transportation Costing Model 
@Rummer2005 assumes a default solid volume factor of 0.3 for biomass meaning that
a 100 m³ capacity trailer could hold the equivalent of 30 m³ of solid biomass.

@@

Timber Tracks produces a publication describing forestry equipment hourly rates
@TT2020.  The rate reported for 
#an 8 axle tridem tractor with a b-train trailer was
tandem tractor with a tandem trailer was
`r sprintf("%5.2f",hourly_truck_rate)` \$/h.  We assume that
the tractor-trailer combination used to haul biomass would cost a 
similar amount.

@@

>>>>

---
title: Spatial and temporal variability in economic availability of residual biomass
  from timber harvest in west-central Alberta, Canada (1990--2015)
author:
- Glen W. Armstrong:
    institute: rr
    correspondence: 'yes'
    email: gwa@ualberta.ca
- Jay A. Anderson:
    email: jaya@ualberta.ca
    institute: rees
- Alex W. McPhee:
    email: awmcphee@ualberta.ca
    institute: rees
institute:
- rr: Department of Renewable Resources, University of Alberta, Edmonton AB  T6G 2H1,
    Canada
- rees: Department of Resource Economics and Environmental Sociology, University of
    Alberta, Edmonton AB  T6G 2H1, Canada
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    number_sections: yes
    df_print: paged
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
  bookdown::pdf_document2: default
  pdf_document:
    includes:
      in_header: preamble.tex
    number_sections: yes
    latex_engine: xelatex
    keep_tex: yes
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
  word_document:
    number_sections: yes
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
  bookdown::html_document2: default
header-includes:
- \usepackage[left]{lineno}
- \usepackage{setspace}
- \linenumbers
geometry: left=1.5in,right=0.5in,top=1in,bottom=1in
csl: biomass-and-bioenergy
bibliography: historicalResidueSupply.bib
---

\raggedright

@@

```{r libraries, echo=FALSE, message=FALSE}
setwd("G:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
#setwd("/mnt/g/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
library(tidyverse)
library(ggthemes)
library(gridExtra)
library(viridis)
library(skimr)
library(scales)
library(DescTools)
library(zoo)
library(ezknitr)
library(magick)
library(kableExtra)
library(gt)
library(glue)
library(quantmod)
library(bookdown)
```

\doublespacing

@@

@@

Biofuels are becoming an increasingly important source of energy because
of their potential to reduce dependency on fossil fuels, thereby
reducing greenhouse gas emissions (IEA Bioenergy 2013). Moreover, second
generation biofuel provides the benefit of avoiding competition with
food production by using lignocellulosic biomass -- often in the form of
agriculture and forest residues -- as feedstock (source?). Forest
residues are generated as by-products of conventional forest harvesting
operations, and include the tree tops, branches and non-merchantable
logs. These residues are typically left behind in slash piles, which are
burned during winter to mitigate against the risk of wildfires. Canada
has a developed forestry sector that produces approximately 20 million
t\[oven dried\]yr^-1^ of forest residues (Dymond et al. 2010).

Despite the potential to produce biofuels from forest residues, such use
is limited to a few highly subsidized pilot projects. Indeed, several
operational and economic challenges hinder the utilization of forest
residues. These challenges are related to: (1) capital investments in
biorefineries, (2) geographical distance between feedstock and markets,
and (3) concerns surrounding feedstock availability (Shabani et al.
2013). The purpose of our study is to explore concerns surrounding this
third challenge -- feedstock availability -- by evaluating annual
variability in commercial stemwood harvest in the province of Alberta,
Canada.

A number of studies have shown that collection and transportation costs
are arguably the biggest constraint to widespread use of forest residues
in bioenergy systems (Ralevic et al 2003, Wood and Layzell 2003, Kumar
et al. 2003, Rummer 2007, Aulakh 2008). Availability of forest residues
for a biorefinery largely depend on commercial harvesting operations and
the production of traditional forest products. Fluctuations in the
volume of stemwood harvested for forest products could have a direct
impact on residues available for biofuels. Major factors, such as the
2008 housing crisis and the ongoing softwood lumber dispute with the
United States have resulted in sawmill closures and reductions in forest
harvesting activities (De Avillez 2014, Spelter 2009). This relationship
between volatile sawlog harvests and subsequent residue variability is
explored in detail by Niquidet and Friesen (2014), who model forest
residue supply in Alberta as a direct function of lumber prices. Given
that biomass feedstock accounts for 40 to 60% of a biorefinery's total
costs (Stephen et al. 2012), reliable feedstock supplies and costs over
the life of the biorefinery are important factors to consider when
deciding where to locate the biorefinery.

Though past studies regarding biorefinery locations have greatly
improved our understanding of feedstock availability and constraints,
one key area that has largely been omitted is considerations of
feedstock supply variability over time. Commercial stemwood harvests
vary year-to-year, largely based on market conditions for forest
products. Since a forest-residue based biorefinery is dependent on
stemwood harvesting for feedstock, any variability in harvesting
directly influences the volume of residues that would be available.
Therefore, considering the variability associated with feedstocks in
different locations could be important in choosing the optimal site for
locating a biorefinery

Although not related to a biorefinery siting decision, some studies have
considered variable feedstock supplies when investigating whether and
when to expand the production of bioenergy from burning residues
(Bolkesjo et al. 2006; Buongiorno et al. 2011; Galik et al. 2009;
Moiseyev et al. 2011). Other studies have explored the context of
variable supplies when making decisions about capital investments in
forest residue-based biorefineries (Cambero et al 2015, Chen and Fay
2011, Papapostolou and Kondili 2011). Variable feedstock supplies are
also shown to affect capital investment decisions related to: investing
in a residue preprocessing facility (Chen and Fay 2011), types of
storage infrastructure for biorefinery feedstocks (Papapostolou and
Kondili 2011), as well as whether to invest in biofuels versus
bioenergy, and when to invest in processing facilities (Cambero et al.
2015).

There has been a number of studies conducted regarding siting decisions
for forest residue-based biorefineries -- see Johnston et al (2012) for
a review of these studies -- however, we are aware of no study that
considered stemwood harvest variability and its subsequent effect on
forest residue variability. Therefore, our study contributes to the
literature by considering variability in forest residue feedstock
supplies. More specifically, we analyze how this variability affects the
cost of delivering residue, and then use these cost estimates to inform
the decision on where to locate a biorefinery.

We consider co-locating the biorefinery adjacent to either a sawmill or
a pulpmill. Co-locating with existing facilities and making use of
established infrastructure and transportation systems can reduce many of
the costs associated with establishing a new biorefinery (Browne et al.,
2012; Saddler et al., 2012; Benjamin et al., 2009; Towers, et al.,
2007). The decision to co-locate a biorefinery near a sawmill or
pulpmill should be informed by how stable the respective residue
supplies are. The facility with a more stable stemwood harvest would
generate a more stable supply of forest residues that could be collected
from the nearby harvested areas.

The objectives of our study are ...... Our approach incorporates
variable feedstock availability into decision making, and our results
provide important information for assessing the potential for a forest
residue-based biofuel industry in Alberta.

The remainder of the paper is as follows: The next section describes
.......

**OTHER POTENTIALLY USEFUL TIDBITS...**

The need to go outside its normal delivery zone in bad years is similar
to a scenario in the agriculture residue feedstock supply literature,
called the the "derisked" scenario, whereby a buffer area outside the
normal supply zone is contracted to provide residues during years when
residue supply is low (Golecha and Gan 2016).

Residue variability is an important source of risk for a forest
residue-based biorefinery. Being dependent upon stemwood harvests for
the generation of residues can lead to cost variability over the long
operating life of a biorefinery, making it is important to have a
comprehensive feedstock risk management program. If the conversion
technology allows it, biorefineries could consider creating a portfolio
of different sources of residues. Such a portfolio approach might
consider locating the biorefinery where it could also access
agricultural residues during periods when availability of forest
residues is low. We plan to investigate feedstock portfolios in future
research.

Modern biorefineries can expect to breakeven financially with delivered
biomass costs of approximately \$80 (US\$60) t\[oven dried\]^-1^
(Steeper 2021) . Studies have shown that forest policy reforms could
reduce delivered forest residue costs by incenting plantations of
fast-growing tree species on lands close to pulpmill locations
(Shooshtarian et al. 2021, Anderson et al. 2012). Other important
factors will be advances in biorefinery technology and increases in
biofuel prices. In Canada, governments impact biofuel prices by setting
renewable fuel mandates and paying subsidies for biofuel production
(Campbell et al., 2016). The continuation of these mandates and
subsidies will likely be an important factor for the emergence of a
second-generation ("advanced") biofuel sector. Also, the Government of
Canada\'s Clean Fuel Standard, which is proposed to come into force in
2022 (Government of Canada, 2020), is expected to improve the economics
of advanced biofuel production.
`

@@

We describe here our study area, data sources, and method.

We use the province of Alberta in western Canada as our study area.
It occupies 661 848 km^2^.  Its southern boundary is 49°N latitude,
northern boundary is 60°N latitude, eastern boundary is 110°W
longitude, and its western boundary is defined by 120°W longitude and
the Great Divide of the Rocky Mountains (Figure 1).

[ Figure 1 about here ]

The southeastern part of the province is largely privately owned
agricultural land, but the northern and western parts are largely
publicly owned forested land.  There is a substantial area of
privately owned agricultural land in the Peace River Country of
northwestern Alberta.  There are also substantial areas in National
Parks along the Rocky Mountains and in the far north of the province.
The province calls the area of mostly publicly owned forested land
(excluding those National Parks) as the Green Area, and the area of
mostly privately owned agricultural land as the White Area.  About 90%
of the harvested timber volume in Alberta comes from the Green Area.

For administrative purposes, the Green Area is divided into Forest
Management Units (FMUs).  The province also enters into long term
agreements with forest products companies giving the companies the
rights to timber harvest and the responsibility to manage the forest
sustainably.  These agreements are called Forest Management
Agreements, and the areas under agreement are called Forest Management
Agreement Areas (FMAAs).  For the most part, the boundaries of the
FMAAs are now coincident with FMU boundaries, with the exception of
the Alberta-Pacific FMAA in eastern Alberta comprises 12 FMUs.  Maps
of the FMUs @AlbertaFMU and the FMAAs @AlbertaFMA can be found on the
Government of Alberta website.  The FMU and FMAA boundaries are
important to this study as they are used to look up average piece size
and utilization standards.  We discuss piece sizes and utilization
standards later.

Our focus for this study is the forest near Hinton, Edson, and
Whitecourt in west-central Alberta.  These towns were chosen as they
host several forest products companies and are located near several
active FMAAs.  We assume that a residual forest biomass processing
plant would be located near an existing forest products mill as the
forest road network has been developed to supply those mills. Locating
a forest residue processing plant near a forest products mill would
also allow for convenient access to mill residues.

The FMAAs which supply most of the timber to mills  near our
three locations are West Fraser Mills Ltd. (Hinton), West Fraser Mills
Ltd. (Edson), Weyerhaeuser Company Limited (Pembina Timberlands),
Millar Western, Blue Ridge Lumber Inc., and ANC Timber Ltd.

We used the Weyerhaeuser oriented
strand board plant in Edson, the Millar Western Forest Products
Ltd. pulp mill in Whitecourt, and the West Fraser pulp mill in Hinton
as the locations for the hypothetical forest residue processing
facilities.

As well as considering locations for the hypothetical plants, we
considered 3 different plant capacities defined by annual biomass
requirements: 200 Gg/a corresponding to a small wood pellet mill, 400
Gg/a corresponding to a large pellet mill or a small ethanol plant,
and 800 Gg/a corresponding to a large ethanol plant.

@@

We used a number of publicly available datasets to create the
township-level summary data used in our analysis.  Most of the data
are freely available on the World Wide Web (WWW). Much of the data
processing required for this study was manipulation of spatial data.
This was done using the QGIS geographic information system software
@QGISorg2021 and add-ons.  All national level data were clipped to the
Alberta provincial boundary for further analysis.

The datasets are  introduced here.  Their use will be discussed
in detail later.

@@

Altalis manages much of the spatial data created by or for the
Government of Alberta.  We use several coverages from their base
features database product @Altalis2021: the provincial, Green/White
area, forest management unit (FMU), forest management agreement area
(FMAA), Alberta Township Survey System township, and town boundaries.

The Alberta Township Survey System (ATS) is particularly important to
this study as all the data used for this study are aggregated or
disaggregated to the township level.  ATS provides the basis for legal
land descriptions in Alberta.  The typical township is roughly square
(9.778 km by 9.716 km, 9 500 ha) and is indexed by township (1--126),
range (1--30), and meridian (4--6).  The townships immediately
east of Alberta's western boundary, and immediately east of the 5^th^
and 6^th^ meridians (corresponding to 114°W and 118°W longitude,
respectively) are partial townships and therefore deviate from the
roughly square shape.  ATS is described in detail by the Alberta Land
Surveyor's Association @ALSA2021.  There are 7 237 townships in
Alberta, allowing for spatial resolution appropriate for regional or
provincial level analyses.

@@

Alberta Geospatial Services is operated by the Alberta Ministry of
Environment and Parks @Alberta2021a.  We obtained road access
coverages from here.  A road access coverage is also available from
Altalis but it does not have as detailed information on road class.

@@

The National Forest Information System (NFIS) provides a set
of spatial data layers representing forest properties at a 250 m
resolution for all of Canada @Beaudoin2017.  We used the merchantable
volume, stand age, and species composition attributes from these
data layers.

Another product of NFIS is a set of spatial datasets derived from
satellite data @NFIS2021b.  The dataset used for this study is the
harvest year data set for the years 1985-2015.  The dataset is
described in detail by Hermosilla *et al.*  @Hermosilla2016.  It
consists of a raster for Canada at 30 m resolution indicating the year
of harvest for each cell.

The NFIS also provides a web-based individual tree biomass calculator
@NFIS2021.  It is used to calculate stem wood, stem bark, branch, and
foliage biomass for a tree based on province, ecozone, species,
diameter at breast height (DBH) and height.  It is possible to upload
a file representing many trees with different combinations of these
input parameters and download a file containing the calculated results
for all the trees.

@@

Ung *et al.*  @Ung2013 describe the tree taper models 
developed for Canada.  A valuable contribution related to this is the WWW-based
wood volume calculator that uses those models @NRCan2015.  We use the
taper models developed by Ung *et al.* to calculate merchantable
volumes and the amount of residual biomass available from cutblocks in
each of the townships in each year.

@@

The national forestry database @CCFM2021 was used to summarize the 
timber harvest volume and harvest area from
Alberta for the years 1990-2015. This information is used to adjust the volumes
obtained from the 250 m resolution inventory to better reflect average harvest
volumes observed in Alberta.

@@

The Alberta government requires FMA holders to develop a forest management
plan for the FMAA.
The forest management plans were used to get information on utilization 
standards and average piece size of harvested trees for the FMAA @Alberta2021.

### Agriculture and Agri-Food Canada

The boundaries of Canada's terrestrial ecozones can be found on
Agriculture and Agri-Food Canada's website @Canada2017.  Ecozone is an
input to the individualt tree biomass calculator @NFIS2021.

# Data aggregation and modeling

The primary objective of this paper was to quantify the spatial and
temporal variability of the costs associated with recovering the
residual biomass left on forest cutblocks following timber harvest.  A
driving assumption for this model was that residue recovery is
opportunistic.  Loggers make the decision about what and where to
harvest, and the biomass recovery occurs at some time following timber
harvest.  The decision to harvest is made independently of any
knowledge of the value of residue as biomass.  The biomass gleaners
provide a service to the loggers as the residue would be piled and
burned if it was not recovered as biomass.

Every township will have associated with it a cycle time (and cost)
required to transport harvest residue from the township to each of our
3 plant locations, a measure of residual biomass for each of the 26
years in the study period. For each plant location and year, the model
looks at all the townships in the province that have a positive
residual volume, and begins the creation of cumulative residual
biomass curves starting with the lowest cycle time first.  These lists
of cycle time and cumulative biomass provide the data for the marginal
cycle time and cost curves.

This is a greedy algorithm.  It ignores the cost of moving equipment
from one block to another, but we believe it still provides useful
information.

@@

The provincial road network was sorted into four classes: primary
highways, other paved roads, two-lane gravel roads, and one-lane
gravel or dirt roads.  Loaded and unloaded travel speeds were assumed
for each road class, according to Table 1. These travel speeds are
based on those assumed by FPInnovations for the Dawson Creek Timber
Supply Area in the neighbouring province of British Columbia
@Blackburn2020.

The harmonic mean of loaded and unloaded speeds was calculated as 
$$
\bar{s_h}=\left(\frac{s_l^{-1} + s_u^{-1}}{2}\right)^{-1} 
$$ 
where $s_l$ is the loaded speed and $s_u$ is the unloaded speed.  Use
of harmonic mean allows us to calculate the correct cycle time: the
time required to travel from the mill, to the collection site at the
cutblock, and back to the mill.  This provides the basis for our
estimates of haul cost.

Figure 3 illustrates the road network in Alberta and estimates of
cycle time based on the average of loaded and unloaded speeds
from Table 1. The road network in the White Area of the province is
very dense and largely arranged as a grid related to the Alberta
Township Survey System.  Most of the roads in the White Area are
public.  The road network in the Green Area is less dense and
many of the roads are built and maintained by resource industries
(*e.g.* forestry and petroleum & natural gas).  The pattern of the
road network is not as regular as in the White Area.

[ Figure 3 about here]

One-way haul times based on the harmonic mean travel speed were
generated using the QGIS Network Analysis Toolbox 3 plug-in (QNEAT3)
created by Raffler @Raffler2018, specifically its iso-area as
interpolation (from point) algorithm.  In our case, we used the
algorithm to determine the minimum one-way travel time (using the
harmonic mean of loaded and unloaded speeds) along the road
network from each of our centres to each cell of a provincial level
raster at a 200 m resolution.  Off-road travel was assumed to be at 5
km/h.  The underlying algorithm for travel time minimization is that
of Dijkstra @Dijkstra1959.  Cycle times were calculated by multiplying
the one-way haul time by 2, and adding 1 hour for loading and unloading

QGIS zonal statistics were used to calculate the median cycle
time to each township, from each of our three centres.  The 
cycle  time in Figure 2 was limited to 14 hours because, in Alberta,
drivers of commercial vehicles are not permitted to drive
more than 13 hours in a 24-hour period, and spend no more that 14
hours on-duty, including loading and unloading times.

[ Figure 2 about here ]

@@

#merch_vol_adj_factor <- 2.61 # based on NFDP
merch_vol_adj_factor <- 2.77 # based on a wild ass guess
#merch_vol_adj_factor <- 2.38 #based on West Fraser Hinton
odmt_adj_factor <-  0.10

@@

from <- c("CAD")
to <- c("USD")
fx <- getQuote(paste0(from, to, "=X"))

@@

All costs and prices used in this study are in Canadian dollars (\$).
At the time of writing (`r sprintf("%s",fx[1,1])` GMT), the exchange rate between
Canadian (CAD) and US dollars (USD) was `r sprintf("%.4f",fx[1,2])` USD/CAD.

All measures of biomass used in this study are reported in SI units on
a dry matter basis:  1 Mg is equivalent to 1 oven dry metric 
ton.

We assume that the biomass available for collection at a cutblock
includes the tops, branches, and foliage left at a roadside landing
after processing by a stroke delimber.  The delimber is assumed to top
the log at the top diameter specified in the utilization standard
relevant to the harvest area.
Our focus in this study is on the spatial and temporal variation in
biomass collection costs based on residue availability associated with
cycle times.  Following Röser @Roser2013,
<!-- the pre-piling cost was set -->
<!-- to `r sprintf("%5.2f",prepiling_cost)` \$·Mg^-1^ , the comminution -->
<!-- cost (@Blackburn2020a) was set to `r sprintf("%5.2f",comminution_cost)` -->
<!-- \$·Mg^-1^ , and other costs (roads, -->
<!-- supervision, overhead, maintenance, and compliance) were set  -->
<!-- to `r sprintf("%5.2f",other_cost)` \$·Mg^-1^ .   -->
the cost of recovering residues from a roadside stroke-delimber , excluding
transport, is `r sprintf("%5.2f",constant_cost)` \$·Mg^-1^.

We assume that the residue will be transported from the forest using a
live floor chip van with 100 m^3^ capacity.  The USDA Forest Service's
Forest Residue Transportation Costing Model @Rummer2005 assumes a
default solid volume factor of 0.3 for biomass meaning that a 100 m^3^
capacity trailer could hold the equivalent of 30 m^3^ of solid
biomass.

@@

Timber Tracks produces a publication describing forestry equipment hourly rates
@TT2020.  The rate reported for 
<!-- an 8 axle tridem tractor with a b-train trailer was -->
tandem tractor with a tandem trailer was
`r sprintf("%5.2f",hourly_truck_rate)` \$/h.  We assume that
the tractor-trailer combination used to haul biomass would cost a 
similar amount.

@@

The areas harvested by township in 2009 according to NFIS @NFIS2021b
@Hermosilla2016 are shown in relationship to the three plant locations and 
the Green Area in Figure 1.  Similar aggregated data were created for
each of the years 1990--2015. For each township, and
each of the years between 1990 and 2015, we store the area harvested (ha) for
use in our analysis.  We can see from this map that a great deal of timber
harvest took place near our hypothetical plant locations in 2009.

These data were created by using QGIS zonal statistics to sum the area
harvested in each township in each year based on the 30 m resolution
harvest data @NFIS2021b.
